Visualizing Contingency Tables

### 2006-12-12 trunk/vcd/paper/Shadings/JCGS.Rnw
JCGS shadings paper accepted
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@629 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[10pt,a4paper]{article}

@@

@@

\SweaveOpts{engine=R,eps=FALSE}
\setkeys{Gin}{width=0.8\textwidth}

@@

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
saturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

lines(c(-.9, -.55), c(0.525, .525), lty = 2)
lines(c(-.45, -.05), c(0.525, .525), lty = 2)
lines(c(.45, .05), c(0.525, .525), lty = 2)
lines(c(.9, .55), c(0.525, .525), lty = 2)
@

<<shading_HCL,fig=TRUE,echo=FALSE,height=7,width=8>>=
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
@
\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

For illustration, a 3-way and a 4-way table are employed: The first is taken
from a case-control study of smoking and {A}lzheimer's disease published in
\cite{vcd:Salib+Hillier:1997} and re-analyzed using conditional inference techniques
in \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}. It provides data on the smoking behaviour
(no, $<$10, 10--20, $>$20 cigarettes per day), disease status (Alzheimer's, other
dementias, other diagnoses) and gender. The question is whether smoking and disease 
status are conditionally independent given gender. All three statistics
suggested above find evidence for departure from significance: the sum-of-squares
statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$
($p = \Sexpr{pval(alz_sumchisq)}$), the maximum sum-of-squares statistic from
Equation~\ref{eq:maxchisq} is $\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$
($p = \Sexpr{pval(alz_maxchisq)}$), and the double maximum statistic from Equation~\ref{eq:dmax}
is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
The $p$ values are again computed by drawing $\Sexpr{nrep_char}$ samples from the corresponding
permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
shows clearly that the association of smoking and disease is present only in
the group of male patients. The double maximum shading employed allows for
identification of the male heavy smokers as the cells `responsible' for the dependence:
other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
under independence. Interestingly, there seems to be another large residual for
the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
is only significant at 10\% and not at the 1\% level as the other two cells.

>>>>

\documentclass[10pt,a4paper,twoside]{article}
\usepackage{fancyheadings}

@@

\pagestyle{fancy}
\thispagestyle{fancyplain}
\headrulewidth0pt
\lhead[]{}
\chead[]{}
\rhead[]{}
\lfoot[This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America]%
{This is a preprint of an article accepted for publication in \textit{Journal of Computational and Graphical Statistics}.
Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}
\cfoot[]{}
\rfoot[]{}

@@

\SweaveOpts{engine=R,eps=FALSE,echo=FALSE,results=hide}
\setkeys{Gin}{width=0.8\textwidth}

@@

\headrulewidth0pt
\lhead[\thepage]{}
\chead[\textit{Residual-based Shadings for Visualizing (Conditional) Independence}]{\textit{Achim Zeileis, David Meyer, Kurt Hornik}}
\rhead[]{\thepage}
\lfoot[{\small Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}]{{\small Copyright {\copyright} 2007 American Statistical Association, Institute of Mathematical Statistics, and Interface Foundation of North America}}
\cfoot[]{}
\rfoot[]{}

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<shading_HSV,echo=FALSE,results=hide>>=
png(file = "shading_HSV.png", height = 525, width = 600)
## generate colors
hue23 <- hue.slice(2/3, grid.n = 101, plot = FALSE, type = "HSV")
hue0 <- hue.slice(0, grid.n = 101, plot = FALSE, type = "HSV")
saturation <- as.numeric(colnames(hue23))
value <- as.numeric(rownames(hue23))

@@

lines(c(-.9, -.55), c(0.525, .525), lty = 2)
lines(c(-.45, -.05), c(0.525, .525), lty = 2)
lines(c(.45, .05), c(0.525, .525), lty = 2)
lines(c(.9, .55), c(0.525, .525), lty = 2)
dev.off()
@
\includegraphics{shading_HSV}

%% height=7,width=8 for PDF

<<shading_HCL,echo=FALSE,results=hide>>=
png(file = "shading_HCL.png", height = 525, width = 600)
## generate colors
hue260 <- hue.slice(260, grid.n = 101, plot = FALSE)
hue360 <- hue.slice(360, grid.n = 101, plot = FALSE)
mychroma <- as.numeric(colnames(hue260))
luminance <- as.numeric(rownames(hue260))

@@

lines(c(-18.75, -11.25), c(52.5, 67.5), lty = 2)
lines(c(-8.75, -1.25), c(72.5, 87.5), lty = 2)
lines(c(18.75, 11.75), c(52.5, 67.5), lty = 2)
lines(c(8.75, 1.25), c(72.5, 87.5), lty = 2)
dev.off()
@
\includegraphics{shading_HCL}
\caption{Residual-based shadings in HSV (upper) and HCL space (lower).}
\label{fig:shadingHSVHCL}
\end{center}
\end{figure}

@@

For illustration, a 3-way and a 4-way table are employed: The first is taken
from a case-control study of smoking and {A}lzheimer's disease published in
\cite{vcd:Salib+Hillier:1997} and re-analyzed using conditional inference techniques
in \cite{vcd:Hothorn+Hornik+VanDeWiel:2006}. It provides data on the smoking behaviour
(no, $<$10, 10--20, $>$20 cigarettes per day), disease status (Alzheimer's, other
dementias, other diagnoses) and gender. The question is whether smoking and disease 
status are conditionally independent given gender. All three statistics
suggested above find evidence for departure from independence: the sum-of-squares
statistic from Equation~\ref{eq:sumchisq} is $\sum_k X_k^2 = \Sexpr{stat(alz_sumchisq)}$
($p = \Sexpr{pval(alz_sumchisq)}$), the maximum sum-of-squares statistic from
Equation~\ref{eq:maxchisq} is $\max_k X_k^2 = \Sexpr{stat(alz_maxchisq)}$
($p = \Sexpr{pval(alz_maxchisq)}$), and the double maximum statistic from Equation~\ref{eq:dmax}
is $\max_k M_k = \Sexpr{stat(alz_dmax)}$ ($p = \Sexpr{pval(alz_dmax)}$).
The $p$ values are again computed by drawing $\Sexpr{nrep_char}$ samples from the corresponding
permutation distribution. The conditional mosaic plot in Figure~\ref{fig:alz}
shows clearly that the association of smoking and disease is present only in
the group of male patients. The double maximum shading employed allows for
identification of the male heavy smokers as the cells `responsible' for the dependence:
other dementias are more frequent and Alzheimer's disease less frequent in this group than expected
under independence. Interestingly, there seems to be another large residual for
the light smoker group ($<$10 cigarettes) and Alzheimer's disease---however, this
is only significant at 10\% and not at the 1\% level as the other two cells.
